Long-term perspectives of social inequality in mortality among

### 2017-07-02 article.Rmd
convert halfway from Rnw to Rmd

\documentclass[a4paper,11pt,titlepage]{article}

\usepackage[utf8]{inputenc}
\usepackage{graphicx}
\usepackage{amsmath}
\usepackage{amssymb}

%\usepackage{float}

\usepackage{epstopdf}
\usepackage{url}

\usepackage{natbib}

\bibliographystyle{apalike}
%\bibliographystyle{plainnat}

%\usepackage[aboveskip=1pt,labelfont=bf,labelsep=period,justification=raggedright,singlelinecheck=off]{caption}
\newcommand{\R}{{\bf R}}

\usepackage[tablesonly]{endfloat}
%\usepackage[figuresonly]{endfloat}
%\usepackage[tablesonly]{endfloat}

\renewcommand{\efloatseparator}{\vfill}

\renewcommand{\textfraction}{0.05}
\renewcommand{\topfraction}{0.8}
\renewcommand{\bottomfraction}{0.8}
\renewcommand{\floatpagefraction}{0.75}

% patch \efloat@iwrite to use \protected@write
\makeatletter
\renewcommand\efloat@iwrite[1]{%
   \immediate\expandafter\protected@write\csname efloat@post#1\endcsname{}}
\makeatother

\author{Sören Edvinsson\footnote{email: \texttt{soren.edvinsson@umu.se}} and
Göran Broström\footnote{email: \texttt{goran.brostrom@umu.se}}
     \\
Centre for Demographic and Ageing Research ({\sc Cedar})\\
Umeå University, SE-90187 Umeå, Sweden}

\title{Long-term perspectives of social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013\footnote{Paper presented
    at the
    session 'Health and Mortality in Arctic Populations'. The Ninth International
    Congress on Arctic Social Sciences,
    Umeå,
    June 8--12, 2017}}

\date{May 31, 2017}

<<setup, include = FALSE>>=
library(knitr)
opts_chunk$set(engine='R',dev='pdf',fig.path='figs/slide-',tidy=FALSE,fig.height=3.5,fig.width=8,strip.white=TRUE,cache=TRUE,echo=FALSE,message=FALSE,comment=NA,show.signif.stars = FALSE,fig.pos='ht')##$
library(knitr)
library(eha)
library(xtable)
library(skum)
@

<<readalldata>>=
## Load remote data:
con <- url("http://capa.ddb.umu.se/data/vb0.rda")
##load("data/vb0.rda")
load(con)
close(con)
vb0$hisclass[is.na(vb0$hisclass)] <- 4
vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "middle", "worker", "NA"))
##
con <- url("http://capa.ddb.umu.se/data/vb.rda")
##load("data/vb.rda")
load(con)
close(con)
##
con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
##load("data/lisaold.rda")
load(con)
close(con)
##
con <- url("http://capa.ddb.umu.se/data/loTab.rda")
load(con)
close(con)
loTab$period <- factor(loTab$period)
##
vb0Tab <- aggregate(vb0[, c("event", "exposure")],
                   vb0[, c("sex", "age", "period", "cohort",
                          "urban", "civst", "hisclass")],
                   FUN = sum)
vb0Tab$age <- factor(vb0Tab$age)
@

\begin{document}

\maketitle

\begin{abstract}
A long-term perspective on the development of social
inequalities in mortality in the adult and elderly population is taken in
this paper. The area under investigation consists of the Skellefteå and Umeå regions in
the north of Sweden, and the time period is 1851--2013. The outcome is all
cause mortality and cause specific mortality for the categories
cardiovascular diseases and cancers. The analysis of inequality is based on
three social classes, reorganized from {\sc Hisclass}.

The main findings are that a consistent social gradient implying better
survival for classes with more access to economic and other resources, only
become evident in the later half of the 20th century, with lower classes
having the highest mortality. The class patterns in mortality differed
however depending on gender. Women of higher classes were comparatively
favoured in most of the studied periods, while higher class men were
disadvantaged when it comes to survival at least until the middle of the
20th century. The modern pattern of clear socio-economic differences
appeared surprisingly late. Additional and alternative analyses based on education and
income categories confirmed large inequality in mortality
between different groups during the most recent decades. There were no clear signs of qualitative differences depending on whether it was the working-active or the retired population that was analysed, but the differences according to social class was larger in the first-mentioned group. Both cardiovascular diseases and cancers conform with the pattern already observed when it comes to all-cause mortality.

\end{abstract}

\tableofcontents

\newpage

%%{\huge Här ska 'intro.tex' in !!!!!!!!!!!!!!!!!!!!!}

%%\subsection{Combining the data sets and tabulate}

\section{Introduction}

One of the central aspects of survival is social class and access to economic and other resources. Even in present-day welfare societies, social position is a strong determinant when it comes to health and mortality and the impact even seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that “… social conditions have been, are and will continue to be irreducible determinants of health outcomes and therefore deserves their appellation as ‘fundamental causes’ of disease and death” \citep{liph95}. For a long time, the general view has been that socio-economic health inequalities were high in historical societies, probably higher than in modern societies. This is a reasonable assumption since these societies were in most cases characterised by very large differences between groups and classes. Knowing that access to resources provides advantages in all ways, the health advantage of higher classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social inequality in health has passed through different historical phases. According to them, social health differences were comparatively small during the pre-transitional phase when space was a strong determinant for the spread of disease. Differences then increased during the transitional phase when mortality declined and wealthy groups used their resources to gain better health. Finally, health differences decreased again in modern low-mortality settings when instead health-related behaviour became important, resulting in marginal differences. \citet{omran82} state, in the third proposition in his theory of the epidemiological transition, that even if the class differential in mortality was maintained during the transition, the decline set in earlier and was faster among privileged groups.

@@

% This paper is a follow-up of an earlier paper \citep{segb16b}, presented at
% the ESHD Conference in Leuven, Belgium, 2016. Here we make a quick look back
% in time for the Skellefteå region (1851--1900), and we also consider causes
% of death from the early 20th century. The {\sc Lisa} data (Swedish population registers) allow
% us to study the effect of income and education on mortality in a more
% direct way for the period 1990--2005, and to compare it with the results of the class-based analysis.

\section{Aims and questions}

@@

% ? :
% 
% \begin{enumerate}
% \item Has inequality in mortality between social classes increased among the adult and elderly
%   population, primarily benefiting the highest social positions and how is it connected to the development of economic inequality in society?
% 
% %%\item Is the mortality transition initiated by a certain social class?
% 
% \item Are there any gender differences in the effect of social position?
% 
% \item Is social position equally important among the retired population as for those in working age?
% \item Is there a difference between classes in mortality from cardiovascular diseases and cancers respectively?
% \item Do class inequalities in survival correspond to those depending on level of education and income categories for the period 1990 on-wards?
% \end{enumerate}
% 
%

\section{The Skellefteå and Umeå regions}

% Figures~\ref{fig:mapp} and~\ref{fig:SKEUME}
% show the study areas and their position in Sweden.

% \begin{figure}
% \includegraphics[height=19cm]{myfigs/skum}
% \caption{The Skellefteå and Umeå regions.}
% \label{fig:SKEUME}
% \end{figure}

@@

\section{Data and variables}

@@

\subsection{Presence periods}

@@

\subsection{Social class}

@@

\begin{enumerate}
\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals and higher white collar.
\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers, farmers and lower white collar.
\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including farm workers.
%%\item Lower skilled farm workers, HISCLASS 10 and 12, Lower and unskilled farm workers
\end{enumerate}

% The procedure for
% occupations in the Linnaeus database is a little more complicated. The
% occupations in the censuses (NYK) are first transferred to the system
% later in use, SSYK (the Swedish version of ISCO88). Thereafter the
% SSYK, both those transferred from the censuses and those given in the
% population registers 2001-2013, can be re-coded into HISCLASS.

@@

% The corresponding numbers for the age group 40--65 are shown in Figures~\ref{fig:hiscskummyoung} and 
% \ref{fig:hiscskumwyoung}. Slightly lower rates of missing occupation in comparison to Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw} are noted.

% <<nanam, results = 'asis'>>=
% library(xtable)
% ##load("data/vb0.rda")
% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
% x <- with(vb0[vb0$sex == "male", ], table(period, hisclass, useNA = "ifany"))
% x <- round(prop.table(x, 1) * 100)
% xtable(x, caption = "Hisclass by time period, men. Per cent.", label = "tab:nana", digits = 0)
% @
% 
% <<nanaw, results = 'asis'>>=
% ##library(xtable)
% ##load("data/vb0.rda")
% ##vb0$hisclass <- factor(vb0$hisclass, labels = c("elite", "lowMan", "lowWhiteC", "worker"))
% x <- with(vb0[vb0$sex == "female", ], table(period, hisclass, useNA = "ifany"))
% x <- round(prop.table(x, 1) * 100)
% xtable(x, caption = "Hisclass by time period, per cent.", label = "tab:nana", digits = 0)
% @

@@

\subsection{Cause of death}
In the present study, we analyse mortality in the two most common disease
groups in the ages we are investigating---cardiovascular diseases and
cancers. National compulsory reporting of cause of death was introduced in 1911, thus our analysis of the two disease groups starts from that year. Classification systems have however changed. Sweden did not join ICD until 1952, but the historical causes of death in the Poplink database have been coded according to the ICD10 system. In the present study, we have identified the codes relevant for cardiovascular diseases and cancers in the ICD 8, 9 and 10.

@@

\subsection{Urban vs rural residence}
In 19th and early 20th century Sweden (as in other countries), mortality was much higher in urban environments. The unsanitary and overcrowded living conditions as well as the risks for importing infections made the health risks higher in towns and cities. In the present paper, we control for this aspect by distinguishing between urban and rural residence, based on in which parish people lived.

\subsection{Periods for analysis}
We have made separate analyses for the different periods. The century 1851--1950 is divided into four periods of 25 years each. The 19th century periods represent a mainly pre-industrial society dominated by agriculture and with restricted welfare provisions. During the years 1901--1950, industrialisation started and the urban environments increased in population size. The modern welfare state began to develop during this time. For the period 1960 on-wards, there are separate analyses on decades except for the two last periods. Sweden had developed into a wealthy welfare society with low income inequality in international comparison, particularly until the 1990s. A financial crisis struck Sweden in the early 1990s, which among other things increased unemployment and changed parts of the welfare system to being less generous than before. The start of the last period (2008--2013) coincides with the international financial crisis.

\subsection{Income}
Income is based on disposable individual income from the LISA database at Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to 2005.

\subsection{Education}
Education is here analysed only for the period 1990 to 2013. Years of education are given in 3 categories and is based on level of education according to categories in the LISA database at Statistics Sweden. It ranges from basic education (nine years of schooling or less) at the lowest level, secondary education (high school or similar) for the middle group and to academic degree at the highest level.

\section{Models}

@@

Another practical problem is the size of that data set:
It contains no less than 2\,432\,871 records of 236\,011 individuals, of
which 68\,244 are observed to die.
This makes it very inconvenient to use the nowadays so popular semi-parametric
technique known as
\emph{Cox regression} \citep{cox72,gb02}. Instead we use a parametric model
that allows for reduction of the data
to \emph{sufficient statistics} \citep{fisher22}. The idea is to keep all
relevant covariates \emph{categorical}, and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three civil statuses, ten periods, two
urban/rural groups, four hisclasses at age 40,  four hisclasses at age 60,
ten age groups, and 22 cohorts, in all 422\,400 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 24\,326. The data set is thus reduced in
size by 99 per cent.

@@

% We analyse both regions together but perform separate analyses for men and
% women and for the age groups 40--64 and 65--89. The model in the analyses
% include social class, sex and marital status.

% We present the results for the period 2001--2013
% separately. We suggest a heterogeneity index in order to investigate a
% possible homogenisation of mortality between social classes.

@@

% There is some  lack of information in the higher
% ages around 1960--1980 (see Figures~\ref{fig:hiscskumm} and \ref{fig:hiscskumw}), because that kind of 
% information is simply missing to a large extent.

\section{The effect of {\sc Hisclass}, 1851--1950, 1961--2013}

\subsection{All causes of death}

@@

%creating an \emph{heterogeneity index} (a better name is needed:
%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
%Something catchy.).

\subsubsection{Ages 40--64}

%The distribution over time periods.
<<getske4060>>=
cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##
cap0 <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap0 <- "Total hazard by HISCLASS, ages 40--64."
cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom)."
scap12 <- "Total hazard standardised by HISCLASS, ages 40--64."
cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
scap12log <- "Total hazard by HISCLASS, ages 40--64, last years."

@@

% We start by looking at cumulative hazards by {\sc Hisclass} over the last four time
% periods, in order to illustrate the process. They are adjusted for {\tt civil status}. We present separate analyses  for women and men, see
% Figures~\ref{fig:hazplotw} and~\ref{fig:hazplotm}.

%\subsubsection{Women}

@@

% 
% 
% <<hazplotw, fig.cap = cap1w, fig.scap = scap1w, fig.height = 7>>=
% source("R/plotHaz.glm.R")
% 
% plotHaz.glm(resw[-(1:6)], main = labb, col = scol, layout = c(2, 2))
% @
% 
% <<hazplotm, fig.cap = cap1m, fig.scap = scap1m, fig.height = 7>>=
% plotHaz.glm(resm[-(1:6)], main = labb, col = scol, layout = c(2, 2))
% @
% 
% From Figures~\ref{fig:hazplotw} and \ref{fig:hazplotm} it is obvious that
% the idea of proportional hazards is not reasonable. We therefore use another approach.

%The "TMR", that is, the summed hazards over the age interval 40--64 is shown in
%Figure~\ref{fig:cumsw}.

@@

%%<<cumsm, fig.cap = "Total haz of deaths in ages 40--64 per individual by HISCLASS and decade, men.">>=
%%@

%% HHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHHH

@@

% \subsubsection{The effect of income and education, 1986--2013}
% 
% A common alternative to a class analysis is to use income or education for investigating socio-economic health differences. 
% Comparing different approaches are furthermore valuable in order to understand why some are advantaged and others 
% disadvantaged. For the time being, we have to restrict such a comparative analysis to the last decades.
% 
% The analysis for the period 1990--2005 is based on a separate analysis with 
% income and education as explanatory variables and with controls of period 
% and marital status.
% 
% <<lissyoung>>=
% library(eha)
% library(xtable)
% liss4064 <- loTab[loTab$age < 64.5, ]
% con <- url("http://capa.ddb.umu.se/data/lisaold.rda")
% ##load("data/lisaold.rda")
% load(con)
% close(con)
% @
% 
% Both education and income had a strong, statistically significant, effect on survival, as expected. 
% 
% \paragraph{Education, 1990--2013}
% 
% <<educ>>=
% 
% educ <- lisaold[!(lisaold$period %in% c("1986-1987", "1988-1989")), ]
% educ$period <- factor(educ$period)
% educ <- educ[!is.na(educ$educ), ]
% levels(educ$educ) <- c("low", "low", "middle", "middle", "middle", "high", "high")
% ##educ$educ <- relevel(educ$educ, ref = "middle")
% educ$civst <- relevel(educ$civst, ref = "married")
% fitm <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
%              data = educ[educ$sex == "male", ])
% fitf <- coxph(Surv(enter, exit, event) ~ strata(educ) + civst + urban,
%              data = educ[educ$sex == "female", ])
% cols <- c("red", "darkgreen", "black")
% @
% 
% <<ploteduc, fig.cap="Survival by education and sex, 1990--2013.">>=
% oldpar <- par(mfrow = c(1, 2))
% 
% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
% 
% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(educ$educ), col = cols, lty = 1, lwd = 1.5)
% 
% par(oldpar)
% @
% 
% Those with the two lowest education level hade the highest mortality, while those with highest level of education had by far the best survival. The differences must be considered large, and it follows almost a continuous improvement according to increasing levels. 
% 
% \paragraph{Income, 1986--2005}
% 
% <<femeduc>>=
% income <- lisaold[!is.na(lisaold$income), ]
% income$period <- factor(income$period)
% ##income <- income[!is.na(income$income), ]
% ##levels(income$income) <- c("low", "low", "middle", "middle", "middle", "high", "high")
% ##income$income <- relevel(income$income, ref = "low")
% income$civst <- relevel(income$civst, ref = "married")
% fitm <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
%              data = income[income$sex == "male", ])
% fitf <- coxph(Surv(enter, exit, event) ~ strata(income) + civst + urban,
%              data = income[income$sex == "female", ])
% cols <- c("darkgreen", "red", "blue", "black")
% 
% @
% 
% <<plotincome, fig.cap="Survival by income and sex, 1986--2005.">>=
% oldpar <- par(mfrow = c(1, 2))
% 
% plot(survfit(fitm, start.time = 40), mark.time = FALSE, main = "Men",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
% 
% plot(survfit(fitf, start.time = 40), mark.time = FALSE, main = "Women",
%      col = cols, lwd = 1.5, axes = FALSE, xlab = "Age", ylab = "Surviving fraction")
% axis(1)
% axis(2, at = c(0, 0.5, 1.0), las = 1)
% abline(h = 0.5, lty = 3)
% box()
% legend("bottomleft", legend = levels(income$income), col = cols, lty = 1, lwd = 1.5)
% 
% par(oldpar)
% @
% 
% Higher incomes result in lower mortality. This is also what could be expected, 
% and conforms to what we know of social health differences in present-day Sweden. 
% Notice however that we cannot establish a causal relationship between 
% income and mortality. Bad health, for instance, may be a confounder.
% 
% %%The statistical significance of the covariates are shown in Table~\ref{tab:liswom}.
% %%<<liswom, results = 'asis'>>=
% %%fitw1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "female", ])
% %%##cat("Women:\n")
% %%drw <- drop1(fitw1, test = "Chisq")
% %%xtable(drw, caption = "Effect of covariates for women 40-64, 1990-2005.", label = "tab:liswom")
% %##round(summary(fitw1)$coef[-(1:13), ], 4)
% %@
% 
% %%The estimated effect sizes for women are shown in Figure \ref{fig:liswomfig}.
% 
% %%<<liswomfig, fig.cap = "Effect sizes, women 40-64, 1990-2005.", fig.height = 4>>=
% %%source("R/plotWEduc.R")
% %%plotWEduc(fitw1)
% %%@
% 
% 
% %%And for men:
% 
% %%<<lismen, results='asis'>>=
% %%fitm1 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss4064[liss4064$sex == "male", ])
% %%##cat("\nMen:\n")
% %%drm <- drop1(fitm1, test = "Chisq")
% %%xtable(drm, caption = "Effect of covariates for men 40-64, 1990-2005.")
% %%##round(summary(fitm1)$coef[-(1:13), ], 4)
% %%dcol <- c("blue", "darkgreen", "darkgreen", "red")
% %%@
% 
% %%<<lismanfig, fig.cap = "Effect sizes, men 40-64, 1990-2005.", fig.height = 4>>=
% %%#source("R/plotWEduc.R")
% %%plotWEduc(fitm1)
% %%@
% 
% % 
% % <<plotliss1, fig.cap = "Cumulative hazards, ages 40--64, 2002--2013.", fig.height = 4>>=
% % #oldpar <- par(mfrow = c(1, 2))
% % #plot(fitw1, fn = "cum", col = dcol, main = "Women")
% % #plot(fitm1, fn = "cum", col = dcol, main = "Men")
% % #par(oldpar)
% % ##summary(fitw1)
% % ##summary(fitm1)
% % @
% % 
% % %See Table~\ref{tab:tablisswom1} the PH regression for women.
% % <<tablisswom1, results = "asis">>=
% % #fitwo1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
% % #                 data = liss4064[liss4064$sex == "female", ])
% % #dr <- drop1(fitwo1, test = "Chisq")
% % #ltx(fitwo1, dr = dr, caption = "PH regression, women aged 40--64, 2002--2013.",
% % #    label = "tab:tablisswom1")
% % @
% % 
% % %See Table~\ref{tab:tablissman1} the PH regression for men.
% % <<tablissman1, results = "asis">>=
% % #fitma1 <- coxreg(Surv(enter, exit, event) ~ DISPINK + CIV + SUN2000NIVA,
% % #                 data = liss4064[liss4064$sex == "male", ])
% % #dr <- drop1(fitma1, test = "Chisq")
% % #ltx(fitma1, dr = dr, caption = "PH regression, men aged 40--64, 2002--2013.",
% %  #   label = "tab:tablissman1")
% % @
%

\subsubsection{Ages 65--89}

%The distribution over time periods.
<<getske6089>>=
vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
vb6589$age <- factor(vb6589$age) # NOTE:: Change to '6589'!
##vb6589$hisclass <- vb6589$hisclass60
@

@@

% \subsubsection{The effect of income and education, 1990--2005}
% 
% The fact that we find rather unclear associations between HISCLASSES and survival, makes alternative approaches to the question of socio-economic health differences interesting. As for the younger age group, we have included an analysis for the period 1990--2005 with income and education as explanatory variables.
% 
% <<lissyoung2>>=
% library(eha)
% library(xtable)
% ##load("data/loTab.rda")
% ##loTab$period <- factor(loTab$period)
% liss6589 <- loTab[loTab$age > 64.5, ]
% @
% 
% 
% The statistical significance of the covariates are shown in
% Table~\ref{tab:liswom2}
% <<liswom2, results = 'asis'>>=
% fitw2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "female", ])
% ##cat("Women:\n")
% drw2 <- drop1(fitw2, test = "Chisq")
% xtable(drw2, caption = "Effect of covariates for women 65-89, 1990-2005.", label = "tab:liswom2")
% ##round(summary(fitw1)$coef[-(1:13), ], 4)
% @
% 
% The estimated effect sizes are shown in Figure \ref{fig:liswomfig2}
% 
% <<liswomfig2, fig.cap = "Effect sizes, women 65-89, 1990-2005.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitw2)
% @
% 
% 
% And for men:
% 
% <<lismen2, results='asis'>>=
% fitm2 <- glm(event ~ offset(log(exposure)) + as.factor(age) + urban + period + civst + income + educ, family = poisson, data = liss6589[liss6589$sex == "male", ])
% ##cat("\nMen:\n")
% drm2 <- drop1(fitm2, test = "Chisq")
% xtable(drm2, caption = "Effect of covariates for men 65-89, 1990-2005.")
% ##round(summary(fitm1)$coef[-(1:13), ], 4)
% dcol <- c("blue", "darkgreen", "darkgreen", "red")
% @
% 
% <<lismanfig2, fig.cap = "Effect sizes, men 65-89, 1990-2005.", fig.height = 4>>=
% #source("R/plotWEduc.R")
% plotWEduc(fitm2)
% @
% 
% These figures show clear associations both with income and education and for women as well as for men. More years of education resulted in lower mortality. For women the association with additional levels became however smaller at the higher levels, the effect is more visible at the steps above the basic levels. In a similar fashion, increased incomes is associated with improved survival both for men and for women. Apparently, education and income is better to catch up socio-economic differences compared to hisclass, at least for this period.

\subsection{Cardiovascular mortality}

@@

First, a general observation is that the analysis of causes of death and its development is difficult to interpret, partly because the number of cases are fewer but also that the quality and coverage of reported death causes differ. Thus, many deaths during the early 20th got quite unspecific causes reported in the death registers. There was still a substantial part of all deaths where cause was diagnosed as unknown or dying from old age, despite the fact that reporting causes of death had become mandatory. For the period 1901--1925 to 1926--1950, the proportion of deaths from cardiovascular diseases increased from 25 to 43 per cent for the age group 40 and above, thus indicating an increase of these diseases. However, at the same time unknown cause and death of old age decreased from 28 percent to 18 per cent. It is quite reasonable to suspect that many cardiovascular deaths were disguised in unspecific categories and that the increase partly is explained by improvements in how deaths were diagnosed.
<<causes41>>=
cvb <- vb0
cvb$event <- cvb$heart ## Can this be done better?
cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
cvb$period <- factor(cvb$period)

@@

%creating an \emph{heterogeneity index} (a better name is needed:
%\emph{health disparity index}, \emph{Social Inequality Mortality Index, (SIMI)}?
%Something catchy.).

\subsubsection{Ages 40--64}

>>>>

---
title: "Long-term perspectives of social inequality in mortality among
  adults and elderly in northern Sweden 1851--2013"
author: "Sören Edvinsson and Göran Broström"
date: "28 juni 2017"
output:
  bookdown::html_document2:
    toc: yes
    toc_depth: 2
    toc_float: true
    number_sections: true
bibliography: ["mybib.bib"]
biblio-style: "apalike"
link-citations: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Introduction

One of the central aspects of survival is social class and access to economic 
and other resources. Even in present-day welfare societies, social position is a 
strong determinant when it comes to health and mortality and the impact even 
seems to be increasing \citep{kuetal04, frlu07}. It has been suggested that 
"... social conditions have been, are and will continue to be irreducible 
determinants of health outcomes and therefore deserves their appellation as 
'fundamental causes' of disease and death" \citep{liph95}. For a long time, 
the general view has been that socio-economic health inequalities were high in 
historical societies, probably higher than in modern societies. This is a 
reasonable assumption since these societies were in most cases characterised by 
very large differences between groups and classes. Knowing that access to 
resources provides advantages in all ways, the health advantage of higher 
classes ought to be obvious. \citet{ant67} and \citet{smi83} suggest that social 
inequality in health has passed through different historical phases. According 
to them, social health differences were comparatively small during the 
pre-transitional phase when space was a strong determinant for the spread of 
disease. Differences then increased during the transitional phase when mortality 
declined and wealthy groups used their resources to gain better health. Finally, 
health differences decreased again in modern low-mortality settings when instead 
health-related behaviour became important, resulting in marginal differences. 
\citet{omran82} state, in the third proposition in his theory of the 
epidemiological transition, that even if the class differential in mortality was 
maintained during the transition, the decline set in earlier and was faster 
among privileged groups.

@@

# Aims and questions

@@

# The Skellefteå and Umeå regions

@@

# Data and variables

@@

## Presence periods

@@

## Social class

@@

\begin{enumerate}
\item {\tt elite}, {\sc Hisclass} 1 and 2. Higher managers, professionals
and higher white collar. 
\item {\tt middle}, {\sc Hisclass} 3, 4, 5, 6 and 8. Lower managers,
farmers and lower white collar. 
\item {\tt worker}, {\sc Hisclass} 7, 9, 10, 11 and 12. Workers including
farm workers. 
\end{enumerate}

@@

@@

## Cause of death

In the present study, we analyse mortality in the two most common disease
groups in the ages we are investigating---cardiovascular diseases and 
cancers. National compulsory reporting of cause of death was introduced in
1911, thus our analysis of the two disease groups starts from that
year. Classification systems have however changed. Sweden did not join ICD
until 1952, but the historical causes of death in the Poplink database have
been coded according to the ICD10 system. In the present study, we have
identified the codes relevant for cardiovascular diseases and cancers in
the ICD 8, 9 and 10.

@@

## Urban vs rural residence

In 19th and early 20th century Sweden (as in other countries), mortality
was much higher in urban environments. The unsanitary and overcrowded
living conditions as well as the risks for importing infections made the
health risks higher in towns and cities. In the present paper, we control
for this aspect by distinguishing between urban and rural residence, based
on in which parish people lived.

## Periods for analysis

We have made separate analyses for the different periods. The century
1851--1950 is divided into four periods of 25 years each. The 19th century
periods represent a mainly pre-industrial society dominated by agriculture
and with restricted welfare provisions. During the years 1901--1950,
industrialisation started and the urban environments increased in
population size. The modern welfare state began to develop during this
time. For the period 1960 on-wards, there are separate analyses on decades
except for the two last periods. Sweden had developed into a wealthy
welfare society with low income inequality in international comparison,
particularly until the 1990s. A financial crisis struck Sweden in the early
1990s, which among other things increased unemployment and changed parts of
the welfare system to being less generous than before. The start of the
last period (2008--2013) coincides with the international financial
crisis.

## Income

Income is based on disposable individual income from the LISA database at
Statistics Sweden, grouped into quartiles. It is only analysed from 1986 to
2005.

## Education

Education is here analysed only for the period 1990 to 2013. Years of
education are given in 3 categories and is based on level of education
according to categories in the LISA database at Statistics Sweden. It
ranges from basic education (nine years of schooling or less) at the lowest
level, secondary education (high school or similar) for the middle group
and to academic degree at the highest level.

# Models

@@

Another practical problem is the size of that data set:
It contains no less than 2432871 records of 236011 individuals, of
which 68244 are observed to die.
This makes it very inconvenient to use the nowadays so popular semi-parametric
technique known as
\emph{Cox regression} \citep{cox72,gb02}. Instead we use a parametric model
that allows for reduction of the data
to \emph{sufficient statistics} \citep{fisher22}. The idea is to keep all
relevant covariates \emph{categorical}, and then create a table with all
possible combinations of levels of the covariates. For each such
combination, the total amount of exposure and the number of deaths are
calculated.  We have two sexes, three civil statuses, ten periods, two
urban/rural groups, four hisclasses at age 40,  four hisclasses at age 60,
ten age groups, and 22 cohorts, in all 422\,400 possible
combinations. However, in this scheme there are many structural zeroes
(impossible combinations) and some random zeros, so the actual number of
combinations we end up with is 24\,326. The data set is thus reduced in
size by 99 per cent.

@@

@@

# The effect of {\sc Hisclass}, 1851--1950, 1961--2013

## All causes of death

@@

### Ages 40--64

<<getske4060>>=
cap1w <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, women."
scap1w <-  "Cumulative hazards, ages 40--64, women."
cap1m <-  "Cumulative hazards for HISCLASS by time period, ages 40--64, men."
scap1m <-  "Cumulative hazards, ages 40--64, men."
##
cap0 <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom)."
scap0 <- "Total hazard by HISCLASS, ages 40--64."
cap12 <- "Total hazard of dying before age 65 for a 40 year old person standardised by HISCLASS and decade, women (top) and men (bottom)."
scap12 <- "Total hazard standardised by HISCLASS, ages 40--64."
cap12log <- "Total hazard of dying before age 65 for a 40 year old person by HISCLASS and decade, women (top) and men (bottom). The last decades."
scap12log <- "Total hazard by HISCLASS, ages 40--64, last years."

@@

@@

@@

@@

### Ages 65--89

<<getske6089>>=
vb6589 <- vb0Tab[as.numeric(as.character(vb0Tab$age)) >= 64.5, ]
vb6589$age <- factor(vb6589$age) # NOTE:: Change to '6589'!
##vb6589$hisclass <- vb6589$hisclass60
@

@@

## Cardiovascular mortality

@@

First, a general observation is that the analysis of causes of death and
its development is difficult to interpret, partly because the number of
cases are fewer but also that the quality and coverage of reported death
causes differ. Thus, many deaths during the early 20th got quite unspecific
causes reported in the death registers. There was still a substantial part
of all deaths where cause was diagnosed as unknown or dying from old age,
despite the fact that reporting causes of death had become mandatory. For
the period 1901--1925 to 1926--1950, the proportion of deaths from
cardiovascular diseases increased from 25 to 43 per cent for the age group
40 and above, thus indicating an increase of these diseases. However, at
the same time unknown cause and death of old age decreased from 28 percent
to 18 per cent. It is quite reasonable to suspect that many cardiovascular
deaths were disguised in unspecific categories and that the increase partly
is explained by improvements in how deaths were diagnosed.

<<causes41>>=
cvb <- vb0
cvb$event <- cvb$heart ## Can this be done better?
cvb <- cvb[!(cvb$period %in% c("1851-1875", "1876-1900")), ]
cvb$period <- factor(cvb$period)

@@

### Ages 40--64
